---
title: "More Git"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```


# Overview

Here I will cover a few more come Git tasks and show some straight-forward ways to do these tasks

* How to clone someone else's GitHub or GitLab repository
* How to clone your own repository--when you want to make a copy and use that as a template for something new.

<!--
* What are branches, merge conflicts and pull requests?
* Fork or clone? What's the difference?
* How to add Git to an existing RStudio project and get that on GitHub or GitLab.
-->

## Downloading other people's repositories

There are two easy ways to do this. Use the one that seems more logical to you.

Method 1.

1. In a browser, go to the GitHub (or GitLab) repository you want to copy.
2. Copy its url.
3. Open GitHub or GitLab.  
4. If using GitHub, click the `+` in top right and click `import repository`. Paste in the url and give your repo a name.
4. If using GitLab, click `New Project` on right, then `Import Repository` tab, then click `Repo by URL`. Paste in url and give repo a name.
5. Open RStudio and click the project tab in the top right and select, `New Project`. Then select `Version Control` and paste in the url of **your** repository's url.  For example, `https://github.com/<youraccount>/Test`  
6. Add the new repo to GitHub Desktop. Open GitHub Desktop, select File>Add Local Repository and navigate to the folder with the new repository.

Method 2.

Steps 1-4 are the same but you can swap step 5 and 6.

5. Open GitHub Desktop. Select File>Clone Repository. Paste in the repository url that you are copying and tell it where to save the repository.
5. Open RStudio and click the project tab in the top right and select, `New Project`. Then select `Existing Directory` and navigate to the directory where you 
just saved the repo.

You can also clone someone elses repo directly into RStudio or GitHub Desktop and then "Publish" to GitHub or GitLab. I am not going to show that but I show that on this [page](Git-RStudio.html). For GitLab, it will require issuing Git commands from a terminal. Note, in my experience, method 1 or 2 above is the way to avoid Git-misery as a Git beginner.

## Making a copy of your own repository

Let say you want to make a copy of a repository and use it as a template to make something else. And you don't want the history!

1. Make a blank repo on GitHub or GitLab (add that Readme file)
2. Pull that down to your computer
3. Copy the repo you want to copy into the new repo folder but **do not copy the .git folder**. Remember the `.git` folder is hidden.



