---
title: "Week 2: Coding Tips"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```


# Week 2. More Git + Better Coding Practices

This week I will cover more Git topics and basic coding habits that I have learned over the years. These are good habits that will help make your code less buggy and easier to extend.

### Agenda

Git, GitHub, GitLab

* How to clone someone else's GitHub or GitLab repository
* How to clone your own repository--when you want to make a copy and use that as a template for something new.
* What are branches, merge conflicts and pull requests?
* Fork or clone? What's the difference?
* How to add Git to an existing RStudio project and get that on GitHub or GitLab.

Coding Tips

* How to organize and plan your code and why adopting an 'object-oriented mindset' will help your code organization (regardless of whether you use object-oriented coding)
* What are namespaces and why you should use `::` to call functions.
* Tips on writing code and functions in R - little things that will make your code better and more robust
* Tips on things to avoid in your R code, i.e. quirks of R that will tend to create bugs


