---
title: "Set-up"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

If you want to follow along with the presentations, you will need R, RStudio, a GitHub account, and GitHub Desktop.

## Get an account on GitHub

If you don't already have an account on GitHub, go to [github.com](https://github.com) and click the "sign up" link near upper right 
of the page.  It is pretty self-explanatory.  Go ahead and get a **free** account.  
The free GitHub account does not have private repositories.  Everything is public and viewable.  You can upgrade (for money) and get private repositories.

## Install GitHub Desktop (Windows and Mac users)

Although you can use Git through RStudio, I am going to show you how to use GitHub Desktop, which is a nice Git/GitHub interface for basic work.

1. Go to [desktop.github.com](https://desktop.github.com/) and install.
2. Open GitHub Desktop, go to 'GitHub Desktop > Preferences' menu.  
3. Under 'Accounts', sign into your GitHub account.
3. Under 'Git', enter your name (or initials) and the email you used for your GitHub account.

Linux users can install Git and interact with GitHub though RStudio.

## Install R and RStudio

* **R:**  Make sure you have a fairly recent version of R.  
Go to [https://cran.r-project.org/](https://cran.r-project.org/) and find the download link for your computer system. Those using a NOAA computer, contact IT to help you update/install.

* **RStudio:** Install the latest version of RStudio.  Get it from [https://www.rstudio.com/products/rstudio/download](https://www.rstudio.com/products/rstudio/download) and install the appropriate one for your OS. Again contact IT to help install if you are on a NOAA computer.

## Set up RStudio to use Git

1. Open RStudio
2. Go to Global Options (from the Tools menu)
3. Click Git/SVN
4. Click Enable version control interface for RStudio projects
5. You will need the location of the git executible. On my mac it is `/usr/bin/git`

Linux users will need to install Git for this to work.  Windows/Mac users: Git comes with GitHub Desktop.

<!--
## Install R packages

From the command line in R Studio

1. **devtools** package.  Enter the following code at the command line (`>`).

```{r get-devtools, eval=FALSE}
install.packages("devtools")  
```

2. Install the **packages** that we are going to be need. I have created a package on GitHub that will install the packages you need.

```{r get-packages, eval=FALSE}
devtools::install_github("RVerse-Tutorials/RWorkflowsetup")
```
-->

## Create a workshop folder

Create a folder/directory on your computer for the workshop materials.  You can create it anywhere you want.  But name it `RWorkflow` just so we all use the same folder name for the workshop.

## Downloading repositories

You will need to download repositories from [RVerse-Tutorials](https://github.com/RVerse-Tutorials) for some of the weeks.  Here is one way to do that. [Watch a video that shows you how to do this](https://youtu.be/b6AaTHFIY7U)

1. In a browser, go to the GitHub repository you want to copy.
2. Click 'Fork' in the upper right corner of the repository.
3. Open RStudio.  Click File > New Project > Version Control
4. Paste in the url of the repository you are copying---the **forked** repository in your GitHub account.  For example, `https://github.com/<youraccount>/Test`  You can leave project name blank.  Replace `<youraccount>` with your GitHub account.
5. Add the new repo to GitHub Desktop. Open GitHub Desktop, select File>Add Local Repository and navigate to the folder with the new repository.

For other ways to download a repository, see this [page](Git-RStudio.html).

## Installing packages from GitHub

Run this code.

```{r get-packages-from-github, eval=FALSE}
devtools::install_github("<github account>/repo name")
```

## (Optional) Packages with C++ code

We will not be doing this, but often you will want to install packages with C++ code.  To do that, you will need Rtools (Windows) or Xcode (Mac)

* **Rtools** Windows users  [Rtools](https://cran.r-project.org/bin/windows/Rtools/) also so you can build packages with C++ code if needed.  See comments here about changing path. [Rtools](http://stat545.com/packages01_system-prep.html#windows-system-prep)

* **Xcode** Mac users Open terminal and type the following command `xcode-select --install`


