---
title: "Set-up"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Everything you'll be using is free. You will need a GitHub (free) account. Sign up here [GitHub](https://github.com/).

To follow along with the material on your own computer, install the following:

* Install [R](https://cran.r-project.org/)
* Install [RStudio](https://www.rstudio.com/products/rstudio/download/)
* Install [GitHub Desktop](https://desktop.github.com/)

To follow along with RStudio Cloud (with a browser) instead of installing programs (or if you following along on a phone or tablet): Sign up for an [RStudio Cloud](https://rstudio.cloud/) account.


# Summary

Summary of the set-up steps. You can also consult this [great online reference](https://happygitwithr.com/).

## Git from RStudio

For the class, I am showing Git within RStudio, but it can be a real headache to get working for beginners. The following set-up is most likely to work. 

1. Get a GitHub or GitLab account
2. Install Git for your operating system
3. Install R (if needed)
4. Install RStudio (if needed)
5. Tell RStudio where git.exe is located
6. Tell Git how to store and get your GitHub/GitLab username and password
7. Do your first Push from the shell to store your username and password.

## Git from GitHub Desktop

I normally work with Git from GitHub Desktop only. It tends not to have the authentication headaches that Git in RStudio has. However, for the class I am going to work with Git from RStudio but I will also show GitHub Desktop. GitHub Desktop doesn't have the authentication headaches that Git in RStudio has.

1. Get a [GitHub](https://github.com/) account
2. Install [GitHub Desktop](https://desktop.github.com/)
3. Open GitHub Desktop, go to Preferences, and login to GitHub.

# Details

### NWFSC staff

**GitHub** If you are using GitHub in a work capacity, as opposed to purely non-work or school, then the repositories you post should be intended for public use (as opposed to just a personal repo). GitHub is officially allowed for a platform for release of products and provides a nice platform for that. I'll show a few examples. There is paperwork to fill out [info](https://docs.google.com/document/d/1KFajoHazgpm55c1wlGeaaq0gRKVoQviAK0yurOQNr8c/view) and [checklist](https://drive.google.com/file/d/0B6s8nwvlch0hZ0VWR01rVmtuUGs/view). Key point is that GitHub is not NOAA-controlled and sensitive information and data that needs to stay on government computers should not be on GitHub. 

**GitHub internal server** If you want to have repositories for work that is not intended for the public or that involve data, code, or reports that cannot be on a non-NOAA computer, then you can use the NWFSC GitHub server. Contact Marcus Nedelmann at NWFSC to get an account. No paperwork needed.

### Non-NOAA participants or those using GitHub purely for non-NOAA work

If you don't already have an account on GitHub, go to [github.com](https://github.com) and click the "sign up" link near upper right of the page.  It is pretty self-explanatory.  Go ahead and get a **free** account. The free GitHub account allows you to have unlimited public and private repositories. 

## Git for your OS

You shouldn't need this...

  * Windows https://gitforwindows.org/ [Read notes here](https://happygitwithr.com/install-git.html#install-git-windows)
  * Mac It should be installed. Open a terminal window and type, `git --version`. If it asks to install XCode, say Yes. [Read notes here](https://happygitwithr.com/install-git.html#macos)

## GitHub Desktop

GitHub Desktop is a nice Git interface for basic work. Despite the name, GitHub Desktop works just as well with GitLab.

1. Go to [desktop.github.com](https://desktop.github.com/) and install.
2. Open GitHub Desktop, go to 'GitHub Desktop > Preferences' menu.  
3. Under 'Accounts', sign into your GitHub account.
4. Under 'Git', enter your name (or initials) and the email you used for your GitHub account.

## R and RStudio

* **R:**  Make sure you have a fairly recent version of R.  
Go to [CRAN](https://cran.r-project.org/) and find the download link for your computer system. Those using a NOAA computer, contact IT to help you update/install.

* **RStudio:** Install the latest version of RStudio.  Get it from [RStudio](https://www.rstudio.com/products/rstudio/download) and install the appropriate one for your operating system (Windows or Mac). Again contact IT to help install if you are on a NOAA computer.

## Set up RStudio to use Git

If you want to use Git within RStudio, do these steps. Note it is not necessary to use Git in RStudio. You can use GitHub Desktop (or any other Git GUI or the terminal command line).

1. Open RStudio
2. Go to Global Options (from the Tools menu)
3. Click Git/SVN
4. Click Enable version control interface for RStudio projects
5. You will need the location of the git executible. On my mac the value in the `Git executible` box is `/usr/bin/git`. On my PC, it is `C:/Program Files/Git/bin/git.exe`. If you install Git for Windows locally, it may be at `C:\Users\your.username\AppData\Local\Programs\Git\bin\git.exe`.

Tell Git how to get your username and password for GitHub

1. Open a terminal window.
2. Type `git config --get credential.helper`
3. If you are on a PC, you should see `manager` and if on a mac, `osxkeychain`
4. If not type,
Mac: `git config --global credential.helper osxkeychain`
Windows: `git config --global credential.helper manager`

There is one more step that comes up on your first push to the remote. You'll need to do your very first Push from the Shell within RStudio so that it knows your username and password.

1. Open up your local repo in RStudio.
2. Make a change to a file and commit the change.
3. Click on the Git tab, click the cog icon and click `Shell...`
4. Type `git push`
5. It should ask for your GitHub or GitLab username and password.

# Help! 

Can't get Git, GitHub and RStudio to connect?

First off, can you get GitHub Desktop to work? If so, I would say just use that for awhile. You don't need to use Git in RStudio.

If you want to use Git in RStudio, read on. You can also consult this [great online resource](https://happygitwithr.com/).

There are two typical problems. 1. Git doesn't know how to ask you for your username and password. 2. Your repo is pointing to the wrong GitHub account. This can happen when you try to copy someone else's account.

### Does RStudio or GitHub Desktop complain about no SSL tokens when you try to Push?

1. Open a terminal window.
2. Paste in this code `git config --global http.sslVerify false`

### Does RStudio complain about needing access to your credentials?

* First, you should have a fairly recent version of RStudio because the Git integration has gotten better.
* If you have a pop-up window, click 'Deny'. Hopefully a terminal window pops up and asks for your GitHub username and password.

### Does RStudio say it cannot find bash.exe?

RStudio doesn't know where `git.exe` is. First, have you installed Git yet?

If you are on Windows and you have not installed Git, install [Git for Windows](https://gitforwindows.org/). Mac users should have it already. Mac It should be installed. Open a terminal window and type, git --version. If it asks to install XCode, say Yes.

Next, tell RStudio where to find `git.exe`. You need to tell it under Tools > Global Options > Git/SVN > Git executible.  On my mac the value in the `Git executible` box is `/usr/bin/git`. On my PC, it is `C:/Program Files/Git/bin/git.exe` but IT installed it for me and installed for all users (so with admin). If you install Git for Windows locally (meaning you install it just for your account instead of for all users on your computer), it should be at `C:\Users\your.username\AppData\Local\Programs\Git\bin\git.exe`


### Does RStudio ask for your username and password every time you try to push?

You need to do a push from the terminal to store your username/password.

1. Open up your project in RStudio and click the Git tab.
2. Click on the cog icon and click `Shell...`
3. Type `git config --get credential.helper` and make sure you see `manager` (or `wincred`) if on Windows and `osxkeychain` if on Mac.
4. If not, look above for the `git` command to add those credential helpers.

### When you try to push with RStudio do you get an error with `askpass`?

RStudio doesn't know how to open a terminal to ask for your username and password. You'll need to push from the shell the first time.

So what's going on is that Git (within RStudio) doesn't know how to ask you for your GitHub username and password.

1. Open up RStudio and make sure your project is showing in the top right with the blue cube with R in it (meaning you have that RStudio project loaded).
2. Make a change to a file in your project and save.
3. Click on the Git tab in the upper right window.
4. Commit that change by clicking on the little box next to the change (in the Git window), adding a comment and clicking Commit.
5. Click on the little cog and select Shell...
6. Type `git config --list`. You should see `credentials.helper=manager` (if on Windows) or `credentials.helper=osxkeychain` (if on Mac). If not type, one of these commands. 
Mac: `git config --global credential.helper osxkeychain` Windows: `git config --global credential.helper manager`.
7. Now Type `git push`. You should be asked for your GitHub username and password.
8. Now make another change to a file, commit the change, and try to Push with the Green arrow.

### Still no luck authenticating? 

Git might have the wrong remote.

1. Open up your project in RStudio.
2. Click the Git tab. Then click on the cog icon and click `Shell...`
3. Type `git remote -v` and see if what is listed looks correct.
4. If not, the command to set the new remote is
`git remote set-url origin URL` Replace `URL` with the URL of your repository!!



<!--
## Install R packages

From the command line in R Studio

1. **devtools** package.  Enter the following code at the command line (`>`).

```{r get-devtools, eval=FALSE}
install.packages("devtools")  
```

2. Install the **packages** that we are going to be need. I have created a package on GitHub that will install the packages you need.

```{r get-packages, eval=FALSE}
devtools::install_github("RVerse-Tutorials/RWorkflowsetup")
```
-->


## Week 2: Installing packages from GitHub

Windows people: You will need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/history.html). 

Run this code to confirm that you can install packages from GitHub:

```{r get-packages-from-github, eval=FALSE}
devtools::install_github("RVerse-Tutorials/TestPackage")
```

To install from an internal GitHub server, like the NWFSC GitHub, use this. Note for the NWFSC GitHub, you need to be on the network since this is an internal (not public) server. Again, replace the dummy variables with the real values.

```{r get-packages-from-gitlab2, eval=FALSE}
devtools::install_github("<github account>/repo name", host="<url of the server>")
```

## (Optional) Packages with C++ code

We will not be doing this, but often you will want to install packages with C++ code.  To do that, you will need Rtools (Windows) or Xcode (Mac)

* **Rtools** Windows users  [Rtools](https://cran.r-project.org/bin/windows/Rtools/) also so you can build packages with C++ code if needed.  See comments here about changing path. [Rtools](http://stat545.com/packages01_system-prep.html#windows-system-prep)

* **Xcode** Mac users Open terminal and type the following command `xcode-select --install`


