---
title: "Set-up"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get an account on GitHub

If you don't already have an account on GitHub, go to [github.com](https://github.com) and click the "sign up" link near upper right 
of the page.  It is pretty self-explanatory.  Go ahead and get a **free** account.  

The free GitHub account does not have private repositories.  Everything is public and viewable.  You can upgrade (for money) and get private repositories.  Or _if you are a student_, you can get private repositories from GitHub for free.  Go to [https://education.github.com/pack](https://education.github.com/pack) to sign up for your free student pack.  You will need to upload proof that you are a student or faculty.

## Install GitHub Desktop (Windows and Mac users)

1. Go to [desktop.github.com](https://desktop.github.com/) and install.
2. Open GitHub Desktop, go to 'GitHub Desktop > Preferences' menu.  
3. Under 'Accounts', sign into your GitHub account.
3. Under 'Git', enter your name (or initials) and the email you used for your GitHub account.

Linux users can install Git and interact with GitHub though RStudio.  GitHub Desktop is not essential for the course, however it is a standard R Workflow that I will be demonstrating.

## Install R and RStudio

* **R:**  Make sure you have the latest version of R.  
Go to [https://cran.r-project.org/](https://cran.r-project.org/) and find the download link for your computer system.

* **RStudio:** Install the latest version of RStudio.  Get it from [https://www.rstudio.com/products/rstudio/download](https://www.rstudio.com/products/rstudio/download) and install the appropriate one for your OS.

## Set up RStudio to use Git

1. Open RStudio
2. Go to Global Options (from the Tools menu)
3. Click Git/SVN
4. Click Enable version control interface for RStudio projects

Linux users will need to install Git for this to work.  Windows/Mac users: Git comes with GitHub Desktop.

## Install R packages

From the command line in R Studio

1. **bookdown:** package.  Enter the following code at the command line (`>`).

```{r get-bd, eval=FALSE}
install.packages("devtools")  
```

2. Install the **packages** that we are going to be need. I have created a package on GitHub that will install the packages you need.

```{r get-packages, eval=FALSE}
devtools::install_github("RVerse-Tutorials/RWorkflowsetup")
```

## Create a workshop folder

Create a folder/directory on your computer for the workshop materials.  You can create it anywhere you want.  But name it `RWorkflow` just so we all use the same folder name for the workshop.

## Downloading repositories

**In case we do not have internet access, I will provide the files on a thumb drive also.**

You will need to download repositories from [RVerse-Tutorials](https://github.com/RVerse-Tutorials) for many of the labs.  Here is one way to do that. [Watch a video that shows you how to do this](https://youtu.be/b6AaTHFIY7U)

1. Go to the repository you want to copy in your browser.
2. Click 'Fork' in the upper right corner of the repository.
3. Open RStudio.  Click File > New Project > Version Control
4. Paste in the url of the repository you are copying---the **forked** repository in your GitHub account.  For example, `https://github.com/<youraccount>/Test-Website`  You can leave project name blank.  Replace `<youraccount>` with your GitHub account.

Another way is to download the repository as a zip file and unzip.  [Watch a video that shows you how to do this](https://youtu.be/bYySjso7vRo)

Another way is to use the function I wrote:

```
library(RWorkflowsetup)
download.repo("Rmarkdown-Tutorial")
```

[Watch a video that shows you how to do this](https://youtu.be/b6AaTHFIY7U)

For other ways to download a repository, see this [page](Git-RStudio.html).

## (Optional) Packages with C++ code

We will not be doing this in this short course, but often you will want to install packages with C++ code.  To do that, you will need Rtools (Windows) or Xcode (Mac)

* **Rtools** Windows users  [Rtools](https://cran.r-project.org/bin/windows/Rtools/) also so you can build packages with C++ code if needed.  See comments here about changing path. [Rtools](http://stat545.com/packages01_system-prep.html#windows-system-prep)

* **Xcode** Mac users Open terminal and type the following command `xcode-select --install`


