---
title: "Week 4. Scrum using GitHub"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```

This is going to be a very brief introduction to Scrum, but hopefully will get you started if you want to try running a Scrum sprint. See the references for more info.

## What is Scrum?

*A lightweight framework for applying Agile methodology for complex projects*

* lightweight = not a lot of rules. easy to set up and follow. easy to adapt.
* framework = a roadmap to follow so you don't have to "make-up" what to do. The Scrum framework is also very popular for a reason.
* Agile management = a flat project management structure that emphasizes individual team members and communications, a working product at all stages, end-user or stakeholder collaboration and feedback, development under the expectation of change
* complex projects = big tasks with lots of different stages, maybe many different team members and different needed skills, could also be something like a R package or manuscript too.




# References

* This is a great 3-hour video course https://learning.oreilly.com/videos/scrum-fundamentals  It is available free through most universities.




