---
title: "Week 6: Advanced package topics"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```
This week I will cover more advanced topics regarding R packages:

* How to use the "check" feature to find problems in your package.
* How to write a vignette for your package.
* How to write print and plot functions for your objects.
* Using pkgdown to make a website for your package.

If/when you want to go into R packaging in more depth, see Hadley Wickham's book [R Packages](http://r-pkgs.had.co.nz/).


## **pkgdown**

[pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) makes it easy to make nice webpages for your packages.


## Depends versus Imports

**@Depends** These packages that will be loaded when your package is loaded. So if you have **gplot2** in @Depends, like above, then the user automatically can use **ggplot2** functions without issuing the command `library(ggplot2)`. 

**@Imports** Are required for the package functions, but the user will not have access to the functions without calling `library(...)`. In your package, you will use `::` to access the functions from the packages on the @Imports line. *Most of your package dependencies will be here.*

To limit the number of headaches that users face when trying to install your package, use as few packages on your @Depends and @Imports lines in DESCRIPTION file as possible. If your package does not need the package to work, then put the package on @Suggests.

I have R packages that are mainly for my personal use. I use the package to make sure I have access to the various packages that I'll be using. So for example, if I am doing work on my sardine papers, I have set of packages that I use. When I issue the command `library(SardineForecast)` a bunch of packages are loaded. This makes it handy for me, but all those dependencies makes it a huge hassle to install the package from GitHub for my collaborators (and even a hassle for me to install from GitHub). Huge Hassle. Invariable one of the 15 packages that I need will itself have a dependency that won't load and then I have to debug that. If I need collaborators, who are on different operating systems and various versions of R to install it, it's a suffer-fest.

For my MARSS package, I have only 3 non-base dependencies in the @Imports line and nothing on the @Depends line besides R. The imports are KFAS, mvtnorm, and nlme. Then on the @Suggests line, I have a bunch of packages that are used in the vignettes. MARSS is easy to install from GitHub (though it is also hosted on CRAN).

**What should you put on your Depends and Imports lines?**

First off, when you are starting, don't worry too much about this. Just add packages that are needed as you work on your functions.

* ALWAYS use `::` to use functions from other packages in your package functions. Seriously. You will save yourself so many headaches down the road by being able to search for `xyzpackage::` to find all that packages functions. Why? Trust me, one day you will want to swap out packages or remove dependencies. Note, this can be a hassle with functions like `ggplot()` which use functions within their calls and you have to use `::` everywhere. Like so
    ```{r eval=FALSE}
    ggplot2::ggplot(df) +
       ggplot2::geom_point(ggplot2::aes(gp, y))
    ```
    Arg. Another example is say a GAM call:
    ```{r eval=FALSE}
    mgcv::gam(a ~ mgcv::s(b), data=df)
    ```
    But this is just for your package functions. In your scripts, you'd probably use a `library()` call.

* Never ever use `library()` in a function! Use `xyzfunction::function`. Sure use `library()` in your scripts, but never in a package function.  When you add a function from a new package to your function, add those packages to @Depends or @Imports in your DESCRIPTION file as you go along.

* Every so often check that you don't have packages on @Depends and @Imports that you don't use. Just do a Edit > Find in Files... search for `xyzpackage::` to find out if you are still using `xyzpackage`.

* How do you know if you forgot a dependency or forgot a `::` somewhere? Two ways:

    1. Do Session > Restart R to close all your loaded packages. Then load your package with `library(yourpackagename)` and try your functions. Things will fail if you have a package in @Imports but forgot a `::` somewhere.
    2. From the Build tab, use More > Check Package. That should show package dependency errors (plus a whole slew of other problems for you to work through).

## Styling your code

You should stick with a uniform style guide to make your code easier to follow. I use the tidyverse style guide with the [**styler** R package](https://www.tidyverse.org/blog/2017/12/styler-1.0.0/). **styler** has an RStudio Addin which does all the work of styling my code for me. Install the package, restart RStudio, and then go to Tools > Addins > Browse Addins. Scroll down to styler, and select the file(s), you want to style.

## Byte Compile

Adding this line to your DESCRIPTION file can really speed up your code. This is one of the advantages of putting your functions in a package. It can actually make your functions faster.

```
ByteCompile: TRUE
```

