---
title: "Week 9: Intro to piping and tidyverse data wrangling"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This week I will give a brief introduction to piping and data wrangling with tidyr and tidyverse verbs. If you want to learn more, there are many resources on the web.  I like [this course book](https://dcl-wrangle.stanford.edu/) as a beginning, [this course](https://cengel.github.io/R-data-wrangling/) goes into more depth, and [R for Data Science](https://r4ds.had.co.nz/) has much more.

Today's goal is to teach you 6 data "verbs" that are easy to use and that will cover many of your data-wrangling needs.

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```


# Piping

You already use the concept of "piping", but you do it with parentheses. Your pipe in this case goes from inside to outside.

```
fun4(fun3(fun2(fun1(x))))
```

`x` is our initial object. We apply `fun1()` to that. Then that result goes into  `fun2()`. The result of that goes into `fun3()`. etc. "Pipe" is referring to the flow of information, the "goes into" part.

This has many problems.

* You will lose track of your parentheses and waste much time sorting out where you lost one of your parentheses.
* You are limited in how many levels you can add without your code being unreadable as it stretches quickly to multiple lines.
* We don't read inside to outside though from our math education, we are used to that (sort of).

Piping is just a different way to write the same thing but left to right like we read (in English).

```
x %>% fun1() %>% fun2() %>% fun3() %>% fun4()
```

Example:
```{r}
1:10 %>% sqrt() %>% mean() %>% round(digits=1) 
```
That's the same as
```{r}
round(mean(sqrt(1:10)),digits=1) 
```

## Upsides

* It is used a lot in the tidyverse so you should be aware of how it works.
* It works well in the tidyverse.
* Can be more readable.
* Makes it easy to avoid intermediate variables (`tmp` and `foo`).

## Downsides

* It's is easy to write code that is hard to understand. It tends to hide where the object on the left is going to in the function on the right.

So you see code like this
```{r}
fun <- function(a=1, b=10){c(max(a), min(b))}
a <- 2:10
a %>% fun(a^2)
```
and you have to go research `fun` to discover that this is `fun(a=a, b=a^2)`.

This kind of mistake is so easy to make. You want say, `rnorm(n=a)`, and accidentally write that in your pipe.
```{r}
a <- 3
a %>% rnorm(n=a)
a %>% rnorm()
```
The left flows into the first function spot, UNLESS you specified that and then it flows to the second or third or whatever is "open".

* You can spend hours trying to solve your piping issue when you could immediately write down code using intermediate values.

* It's really slow.
```{r message=FALSE}
library(microbenchmark)
bm <- microbenchmark(1:10 %>% sqrt() %>% mean() %>% round(digits=1), round(mean(sqrt(1:10)),digits=1) )
ggplot2::autoplot(bm)
```

* You must stay hyper-alert as to the names in the data frame you are working with. You might not have created this data frame. This may be a function you wrote to take any data frame from a user.... Otherwise, bad things like this can happen.
```{r}
fun <- function(x, cyl=1){
x %>% transform(new = mpg^cyl) %>% select(mpg, new)
}
```
Works great!
```{r}
df <- data.frame(mpg=runif(10, 10, 30))
fun(df, cyl=0.5)
```
Why, why? It didn't complain but this is wrong!
```{r}
fun(mtcars[1:5,], cyl=0.5)
```

  
# Tidyverse verbs

Read up on tidyverse verbs [here](https://dplyr.tidyverse.org/).

NOTE: 

* I am using piping because that is how you will see these verbs used in online material. They are just functions and you do not need to use them with piping. 
* They are very slow. For loops are not the place for these functions. `apply()` in base R is much faster.
* Tidyverse is maturing and the semantics across functions is not quite the same. Careful.

## Workflow

```
data.frame %>%
  select columns or filter rows %>%
  summarize or mutate
```

```
data.frame %>%
  group data %>%
  summarize or mutate
```

Verbs that we will learn:

* `select()` select columns
* `filter()` select rows
* `group_by()` apply the function differently within groups.
* `summarize()` apply function over columns -> fewer rows. Like `apply(df,2,function)`
* `mutate()` apply function to columns and make new column. Like `apply(df,1,function)`

What we won't learn: applying a function across a group of columns. See ?across to learn about that.

## Selecting and filtering

```
data.frame %>%
  select or filter
```

### `select()`

Select columns.

```{r}
head(mtcars)
```

```{r}
mtcars %>% select(mpg)
```

```{r}
mtcars %>% select(-mpg, -qsec, -gear, -cyl)
```
```{r}
mtcars %>% select(disp:qsec)
```

```{r}
mtcars %>% select(!disp:qsec)
```
```{r}
df %>% select(where(is.numeric))
```

### `filter()`

Select rows. 
```{r}
mtcars %>% filter(cyl==4)
```
Same as `subset(df, cyl==4)` or `df[df$cyl==4,]`

```{r}
mtcars %>% filter(cyl==4, hp>90)
```
Same as `subset(df, cyl==4 & hp>90)` or `df[df$cyl==4 & df$hp>90,]`.

## `summarize()`

`summarize()` summarizes down columns and leads to fewer rows. **You must tell it the columns to work on.**

```
data.frame %>%
  summarize
```

Apply a function to columns.
```{r}
mtcars %>% 
  summarize(mean.wt=mean(wt))
```
I can add on more functions.
```{r}
mtcars %>% 
  summarize(mean.wt=mean(wt)) %>%
  round(digits=2)
```

Same as
```{r}
round(mean(mtcars$wt), digits=2)
```

You can specify multiple columns. You have to put in `c()` in this case.
```{r}
mtcars %>% 
  summarize(mean.wt=mean(c(mpg, hp)))
```
Same as
```{r}
mean(c(mtcars$mpg, mtcars$hp))
```

Sadly this does NOT work for specifying a range. It computes something but not what you'd think. **This is the  `apply across columns` case.** Read up on that in `?across`.
```{r}
mtcars %>% 
  summarize(mean.wt=mean(mpg:hp))
```
This is computing the mean of this:
```{r}
mtcars[1,"mpg"]:mtcars[1,"hp"]
```
which is
```{r}
mean(mtcars[1,"mpg"]:mtcars[1,"hp"])
```

## `summarize()` by column


`summarize_all()` and `summarize_at()`

Apply to a range of columns or all columns. Note the `_at()` and `_all()` functions have been superseded with `across()` but I find `across()` really cumbersome. I personally would use `select()` and `summarize_all()`.
```{r}
mtcars %>% summarize_at(vars(mpg:hp), mean)
```

```
data.frame %>%
  select columns %>%
  use summarize_all()
```

```{r}
mtcars %>% 
  select(mpg:hp) %>% 
  summarize_all(mean)
```
```{r}
apply(mtcars[,1:4],2,mean)
```

`summarize_all()` is less picky than `apply()`
```{r}
df <- mtcars[,1:3]; df$label="test"
df %>% summarize_all(mean)
```
while `apply()` gives all NA.
```{r}
apply(df,2,mean)
```

Another example.
```
data.frame %>%
  select rows %>%
  select columns %>%
  use summarize_all() verb to apply %>%
  format with round()
```

```{r}
mtcars %>% 
  filter(cyl==4) %>% 
  select(mpg:hp) %>% 
  summarize_all(mean) %>%
  round(digits=2) %>%
  paste(collapse=" -- ")
```

## `summarize()` by group

```
data.frame %>%
  define groups %>%
  apply summarize within groups
```

```{r}
df2 <- data.frame(id=paste0(rep("s",8),1:2),
                  pop=paste0(rep("r",8),rep(1:2,each=4)),
                  Week1=rnorm(8), Week2=rnorm(8), Week3=rnorm(8), Week4=rnorm(8))
head(df2)
```


The `group_by()` function allows you to do more complicated things. Take the mean by pop.
```{r}
df2 %>% group_by(pop) %>% summarize_all(mean)
```

```{r}
apply(df2[df2$pop=="r1",3:6],2,mean)
```

Take the mean for all weeks by pop and id. Notice that I have to specify columns for `summarize()`.
```{r}
df2 %>% 
  group_by(pop, id) %>%
  summarize(mean=mean(c(Week1, Week2, Week3, Week4)))
```

## Combine group and select

Select based on properties. For some reason you have to wrap in `where()`.
```{r}
df2 %>% group_by(pop) %>% 
  select(where(is.numeric)) %>% 
  summarize_all(mean)
```

Select based on name.
```{r}
df2 %>% group_by(pop) %>% 
  select(ends_with("2")) %>% 
  summarize_all(mean)
```

## `mutate()`

Apply a function across rows and return a new column. Like `apply(df,1,fun)`.

```{r}
mtcars %>% 
   select(mpg:wt) %>%
   mutate(wt.hp=wt/hp)
```



You can write custom functions to apply to the groups. 

```{r}
fun <- function(x,y){fit <- lm(y~x); coef(fit)[1]} #return interept
mtcars %>% 
  group_by(carb) %>% 
  summarize(lmm=fun(mpg,hp))
```

Use `mutate()` to add column. **This will not change the number of rows.**
```{r}
mtcars %>% 
  group_by(carb) %>% 
  mutate(lmm=fun(mpg,hp)) %>%
  select(mpg, hp, lmm)
```


### Careful `mutate()` guesses

That works.
```{r}
mtcars %>% 
   select(mpg:wt) %>%
   mutate(mean=sqrt(wt))
```

Yipes! It decided *you must want me to take the mean in the whole column*.
```{r}
mtcars %>% 
   select(mpg:wt) %>%
   mutate(mean=mean(wt))
```

### Careful regarding *across*

It is super easy to try to use this to apply functions across multiple columns. Doesn't work. You need to use `?across` (or `group_by()` with a row factor).

```{r}
mtcars %>% 
   select(mpg:wt) %>%
   mutate(mean1 = (wt+hp)/2, mean2=mean(c(wt,hp)))
```


## Tidyverse verbs are slow

For loops are not the place for these verbs!
```{r message=FALSE}
bm <- microbenchmark(
  apply(mtcars,2,sum), 
  summarize_all(mtcars, sum)
  )
ggplot2::autoplot(bm)
```

```{r message=FALSE}
bm <- microbenchmark(
   mtcars[mtcars$cyl==4,], 
   filter(mtcars, cyl==4),
    subset(mtcars, cyl==4)
   )
ggplot2::autoplot(bm)
```

# Long and wide data frames

Wide is what you need for say `lm()`. You have your response variable in one column and then your predictor variables in other columns.

```{r}
df <- subset(mtcars, select=c(mpg, wt, hp, carb))
df$model <- rownames(mtcars)
head(df)
```
Fit with `lm()`.
```{r}
fit <- lm(mpg ~ wt + hp, data=df)
summary(fit)
```

Long form is what ggplot wants. Usually I have to ponder a bit before I decide how to make my long form data frame. There is not just "one" that you could make from your wide form data frame.

Let's make a plot of the red, green and blue columns. We want the points to be red, green or blue.
```{r}
plot(df$mpg, col="red", ylim=c(0,200))
points(df$wt, col="green")
points(df$hp, col="blue")
```

Same idea but let's plot versus the carburator number.
Let's make a plot of the red, green and blue columns. We want the points to be red, green or blue.
```{r}
plot(df$carb, df$mpg, col="red", ylim=c(0,200))
points(df$carb, df$wt, col="green")
points(df$carb, df$hp, col="blue")
```

Now let's do this in ggplot. What values do we want in the `aes()` function? Those need to be columns. `aes(x=..., y=..., col=....)`
```
data.frame(
   col=name.of.thing.we.plot, 
   y=value.of.thing.we.plot, 
   x=x.value
   )
```

So we'll have a column of x (carb or model or 1:n), a column for our y (mpg,wt,hp) and a column for names ("mpg","wt","hb").
```{r}
df.long <- data.frame()
a <- data.frame(model=df$model, val=df$mpg, name="mpg", carb=df$carb)
df.long <- rbind(df.long, a)
a <- data.frame(model=df$model, val=df$wt, name="wt", carb=df$carb)
df.long <- rbind(df.long, a)
a <- data.frame(model=df$model, val=df$hp, name="hp", carb=df$carb)
df.long <- rbind(df.long, a)
df.long
```

Now we can plot with ggplot.
```{r}
ggplot(df.long, aes(x=model, y=val, col=name)) + geom_point()
```

And we can easily do panels.
```{r}
ggplot(df.long, aes(x=model, y=val, col=name)) + geom_point() +
  facet_wrap(~name)
```

## Pivoting

So creating your long data frames like that would get tidieous and inefficient.
The tidyr package has [functions to make this easier](https://dcl-wrangle.stanford.edu/pivot-basic.html).

### `pivot_longer`

```
pivot_longer(
    data frame name,
    cols = what columns will you stack or pivot,
    names_to = colname for the column with the column names that your stacked,
    values_to = colname for the column with the values
  )
```

```{r}
df.long2 <- 
  pivot_longer(
    df,
    cols = c("mpg", "wt", "hp"),
    names_to = "name",
    values_to = "val"
  )
```

```{r}
df.long2
```

It changed the order our names but that's ok. We can fix that if needed.
```{r}
ggplot(df.long, aes(x=model, y=val, col=name)) + geom_point() +
  facet_wrap(~name)
```

It changed the order our names but that's ok. We can fix that if needed.
```{r}
df.long$name <- factor(df.long$name, levels=c("mpg", "wt", "hp"))
ggplot(df.long, aes(x=model, y=val, col=name)) + geom_point() +
  facet_wrap(~name) +
  theme(axis.text.x=element_blank())
```

Read `?tidyr::pivot_longer` to learn more options but this will be all you need most of the time.

### `pivot_wider`

```
pivot_wider( 
    data frame name,
    names_from = what column has the column names
    values_from = what column has the values
  )
```


```{r}
pivot_wider( 
    df.long,
    names_from = "name", 
    values_from = "val"
  )
```
BTW, you will normally see the function called with piping.

```{r results="hide"}
df.long %>%
  pivot_wider( 
    names_from = "name", 
    values_from = "val"
  )
```


# Comments

I rarely use piping. Why?

* Piping is fine but is super slow. Also I develop packages and the R package checks often through errors with piping or any type of implicit reference to data columns in data frames.  So like `subset(df, x==1)` is implicit because it is implying that there is an `x` column in `df` but that code will actually work if `x` is not in `df`. If you are a package writer, the check will complain about that vagueness.
* I have a hard time following long strings of pipes and lose track of what's going on.

And I rarely use dplyr, tidyr, or tidyverse verbs.

* tidyverse is "maturing" which means the functions keep changing...which means my code rots quickly.
* I often have to do something that takes me a long time to figure out how to do with tidyverse.

Whereas learning ggplot2 has made my plotting much, much faster, I can't say the same with data-wrangling with tidyverse except for a few functions that I use to help out with ggplot.